Towards Practical First-Order Model Counting

### 2024-08-09 doc/paper/paper.tex
Some minor changes

% TODO: consider restructuring the paper (how?)

@@

% TODO: Use FastWFOMC.jl instead, removing the mention of 'privately-obtained'

@@

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in WFOMC:
\begin{enumerate*}[label=(\roman*)]
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{};
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22};
  \item \UFO{} is the input format supported by a private version of
  \textsc{FastWFOMC} obtained directly from the authors.
\end{enumerate*}
Note that the publicly available
version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
of \textsc{FastWFOMC} does not support any cardinality constraints. The notation
we use to refer to each logic is standard in the case of \Ctwo{} and redefined
to be more specific in the case of the other two logics. All three logics are
function-free, and domains are always assumed to be finite. As usual, we
presuppose the \emph{unique name assumption}, which states that two constants
are equal if and only if they are the same constant~\cite{DBLP:books/aw/RN2020}.

@@

@@

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we use
the FOKC algorithm of the original \textsc{Crane} to compile the formula into
three components:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification (explained at the end of this section),
$\mathcal{E}$ is passed to the \FindBaseCases procedure (explained in
\cref{sec:identifying}), which identifies the base cases that need to be
defined.

For each base case $f(\mathbf{x})$, we retrieve the logical formula
$\mathcal{F}(f)$ associated with the function name $f$ and simplify it using the
\Propagate procedure (explained in detail in \cref{sec:simplifying}). We do this
by iterating over all indices of $\mathbf{x}$, where $x_{i}$ is a constant, and
using \Propagate to simplify $\psi$ by assuming that domain $\mathcal{D}(f, i)$
has size $x_{i}$. Finally, on \cref{line:final}, \CompileWithBaseCases recurses
on these simplified formulas and adds the resulting base case equations to
$\mathcal{E}$. \Cref{example:overall} below provides more detail.

@@

% TODO: F missing

% TODO: E is not explained other than that we take the example from X

% TODO: include a high-level overview of first-order knowledge compilation and the meaning of the resulting functions. For the most part, the compilation process that constructs these function definitions can be treated as a black box for the purposes of this work. However, a few high-level details still ought to be mentioned:
% 1. Each algebraic variable ("m", "n", and "l" in the example) corresponds to a domain size.
% 2. \mathcal{D} indicates that "m" denotes the size of \Gamma, "n" denotes the size of \Delta, etc.
% 3. \mathcal{E} can contain any (positive) number of functions.
% 4. Function denoted by "f" represents the solution to the model counting problem.
% 3. The model count of the original formula is computed by evaluating f on the desired domain sizes. For example, f(3, 2) is the model count when |\Gamma| = 3, and |\Delta| = 2.
\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by
  \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}) that
  defines predicate $P$ as a bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Compile. Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we obtain the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, resulting in $g(l, 0) = 1$.
\end{example}

@@

% TODO: The asymptotic analysis of the total number of arithmetic operations
% required to compute the FOMC of some of these formulas was already performed
% in previous work~\cite{DBLP:conf/kr/DilkasB23}.

@@

For the rest of this section, let $\mathcal{E}$ represent the equations returned
by \CompileWithBaseCases. To demonstrate that the base cases identified by
\FindBaseCases are sufficient, we begin with a few observations that stem from
the details of previous
work~\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:conf/kr/DilkasB23} and this work.

>>>>

@@

@@

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in WFOMC:
\begin{enumerate*}[label=(\roman*)]
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{};
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22};
  \item \UFO{} is the input format supported by the most recent implementation
  of
  \textsc{FastWFOMC}\footnote{\url{https://github.com/jan-toth/FastWFOMC.jl}}.
\end{enumerate*}
The notation we use to refer to each logic is standard in the case of \Ctwo{}
and \UFO{}~\cite{tóth2024complexityweightedfirstordermodel} and redefined to be
more specific in the case of \FO{}. All three logics are function-free, and
domains are always assumed to be finite. As usual, we presuppose the
\emph{unique name assumption}, which states that two constants are equal if and
only if they are the same constant~\cite{DBLP:books/aw/RN2020}.

@@

Before describing the main contribution of this work, let us review the key
aspects of FOKC as realised by \textsc{Crane}. The input formula is compiled
into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
$\mathcal{E}$ can contain any number of functions, one of which (denoted by $f$)
represents the solution to the FOMC problem. To compute the FOMC of the formula
for particular domain sizes, $f$ must be evaluated with those domain sizes as
arguments. $\mathcal{D}$ records this correspondence between function arguments
and domains.

@@

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we use
the FOKC algorithm of the original \textsc{Crane} to compile the formula into
the three components: $\mathcal{E}$, $\mathcal{F}$, and $\mathcal{D}$. After
some algebraic simplification (explained at the end of this section),
$\mathcal{E}$ is passed to the \FindBaseCases procedure (explained in
\cref{sec:identifying}), which identifies the base cases that need to be
defined. For each base case $f(\mathbf{x})$, we retrieve the logical formula
$\mathcal{F}(f)$ associated with the function name $f$ and simplify it using the
\Propagate procedure (explained in detail in \cref{sec:simplifying}). We do this
by iterating over all indices of $\mathbf{x}$, where $x_{i}$ is a constant, and
using \Propagate to simplify $\psi$ by assuming that domain $\mathcal{D}(f, i)$
has size $x_{i}$. Finally, on \cref{line:final}, \CompileWithBaseCases recurses
on these simplified formulas and adds the resulting base case equations to
$\mathcal{E}$. \Cref{example:overall} below provides more detail.

@@

\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by
  \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}) that
  defines predicate $P$ as a bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Compile. (We omit the
  definition of $\mathcal{F}$ as the formulas can get a bit verbose.) Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we obtain the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, resulting in $g(l, 0) = 1$.
\end{example}

@@

@@

In the rest of this section, we show that the base cases identified by
\FindBaseCases are sufficient for the algorithm to terminate. Note that
characterising the fine-grained complexity of the solutions found by \Cranetwo{}
or other FOMC algorithms is an emerging area of research partially addressed by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23} and
\citeauthor{tóth2024complexityweightedfirstordermodel}~\shortcite{tóth2024complexityweightedfirstordermodel}.
For the rest of this section, let $\mathcal{E}$ represent the equations returned
by \CompileWithBaseCases. To demonstrate that the base cases identified by
\FindBaseCases are sufficient, we begin with a few observations that stem from
previous~\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:conf/kr/DilkasB23} and this
work.
